\textsc{Atomic modeling of argon}\\ \footnotesize{\textsc{Project 5 --- FYS3150}}

### 2015-12-03 projects/project_5/molecular_dynamics/article/article.tex
diffusion constant done
cell lists explained

    The Lennard-Jones potential approximates the interaction between a pair of
    atoms. In its most common form it reads
    \begin{equation}
        \notag U(r_{ij}) = 4\varepsilon \left[ \left( \frac{\sigma}{r_{ij}}
        \right)^{12} - \left( \frac{\sigma}{r_{ij}} \right)^6 \right],
    \end{equation}
    where $r_{ij}$ is the distance between atom $i$ and atom $j$ and
    $\varepsilon$ is the depth of the potential well. The quantity $\sigma$ is
    the distance for which the potential is zero.  With an expression for the
    potential we can sum over all distinct pairs of atoms and acquire an
    expression for the total potential energy $V$:
    \begin{equation}
        \notag V = \sum^{}_{i>j}U(r_{ij}).
    \end{equation}
    The force exerted on atom $i$ by atom $j$ is then given as the negative
    gradient of the potential between the two:
    \begin{equation}
        \notag \vec{F}(r_{ij}) = -\nabla U(r_{ij}).
    \end{equation}
    The three force-components are given by:
    \begin{align*}
        F_q(r_{ij}) = -\frac{\partial U}{\partial r_{ij}} \frac{\partial
        r_ij}{\partial q_{ij}} \quad (q = x, y, z)
    \end{align*}
    Analytically, this evaluates to
    \begin{align*}
        F_q(r_{i, j}) = 24\varepsilon \left[ \left(
        \frac{2\sigma^{12}}{r_{ij}^{13}}\right) - \left(
        \frac{\sigma^6}{r_{ij}^7}\right)\right] \quad (q = x, y, z).
    \end{align*}
    The brute force way of computing this is summing over each distinct pair of
    atoms, and computing the mutual force between them. This however,
    constitutes a fairly heavy computational load that quickly scales with the
    size of the system. For a total of $n$ atoms in the system, we have a time
    complexity of $\mathcal{O}(n^2)$. Luckily more efficient methods are
    available, most notably is the Verlet list method.

@@

    In order to measure the melting temperature of the system, we can use the
    Einstein relation which states the following:
    \begin{equation}
        \notag
        \langle r^2(t) \rangle = 6Dt.
    \end{equation}
    $D$ is here the diffusion constant, $t$ is time and the \emph{mean square
    displacement} is calculated as
    \begin{equation}
        \notag
        r_i^2(t) = \left| \vec{r}_i(t) - \vec{r}_i(0) \right|^2
    \end{equation}
    for atom $i$. Solving this for $D$ gives yields
    \begin{equation}
        \notag
        D = \frac{\langle r^2(t) \rangle}{6t}.
    \end{equation}
    We can now make set of simulations for varying initial temperature $T$.

\clearpage

>>>>

    The Lennard-Jones potential approximates the interaction between a pair of
    atoms. In its most common form it reads
    \begin{equation}
        \notag U(r_{ij}) = 4\varepsilon \left[ \left( \frac{\sigma}{r_{ij}}
        \right)^{12} - \left( \frac{\sigma}{r_{ij}} \right)^6 \right],
    \end{equation}
    where $r_{ij}$ is the distance between atom $i$ and atom $j$ and
    $\varepsilon$ is the depth of the potential well. The quantity $\sigma$ is
    the distance for which the potential is zero.  With an expression for the
    potential we can sum over all distinct pairs of atoms and acquire an
    expression for the total potential energy $V$:
    \begin{equation}
        \notag V = \sum^{}_{i>j}U(r_{ij}).
    \end{equation}
    The force exerted on atom $i$ by atom $j$ is then given as the negative
    gradient of the potential between the two:
    \begin{equation}
        \notag \vec{F}(r_{ij}) = -\nabla U(r_{ij}).
    \end{equation}
    The three force-components are given by:
    \begin{align*}
        F_q(r_{ij}) = -\frac{\partial U}{\partial r_{ij}} \frac{\partial
        r_ij}{\partial q_{ij}} \quad (q = x, y, z)
    \end{align*}
    Analytically, this evaluates to
    \begin{align*}
        F_q(r_{i, j}) = 24\varepsilon \left[ \left(
        \frac{2\sigma^{12}}{r_{ij}^{13}}\right) - \left(
        \frac{\sigma^6}{r_{ij}^7}\right)\right] \quad (q = x, y, z).
    \end{align*}
    The brute force way of computing this is summing over each distinct pair of
    atoms, and computing the mutual force between them. This however,
    constitutes a fairly heavy computational load that quickly scales with the
    size of the system. For a total of $n$ atoms in the system, we have a time
    complexity of $\mathcal{O}(n^2)$. Luckily more efficient methods are
    available, most notably is the Cell list method.

    \subsubsection{Cell lists}
    \label{ssub:cell_lists}

    The main idea behind cell lists is that at some fixed distance, the force
    induced by the potential between two atoms is negligible and there is
    therefore no need for computing it. If we for each atom $i$ keep track of
    all atoms within this fixed distance, which we will call $r_\mathrm{cut}$,
    at any given time step, we can greatly reduce the number of computations
    needed.

    We divide the domain of our system into \emph{cells}, each with an edge
    length of at least the cut-of radius of the potential. This ensures that if
    we only compute the mutual forces between atoms in the same or adjacent
    cells, then the atoms are never a distance farther than $r_\mathrm{cut}$
    apart. For each integration step we check the validity of the list of
    neighbors for each atom by iterating through each atom in the system, and
    checking whether it has moved a distance larger than $r_\mathrm{cut}$ since
    the last time the list of neighbors was built.

    We additionally need to rescale our potential so that it takes into
    consideration the cut-off radius. We define the scaled potential $U' =
    U(r_{ij}) - U(r_\mathrm{cut})$ when $r_{ij} < r_\mathrm{cut}$ and 0
    otherwise.

    There is a quite subtle consideration which is often overlooked, and that
    is the fact that in order to make use of cell-lists, we need a sufficient
    number of unit cells in each dimensions. Otherwise, there is not enough
    room for our cells.  In the program, if the number of unit cells is less
    than or equal to 5, we employ our old potential. If it is larger, we use
    the new scaled potential with cell lists.

    By implementing cell-lists we reduce the computation of the forces from
    $\mathcal{O}(N^2)$ to $\mathcal{O}(N)$.

    \subsubsection{Choice of parameters}
    \label{ssub:choice_of_parameters}

@@

    In order to measure the melting temperature of the system, we can use the
    Einstein relation which states the following:
    \begin{equation}
        \notag
        \langle r^2(t) \rangle = 6Dt.
    \end{equation}
    $D$ is here the diffusion constant, $t$ is time and the \emph{mean square
    displacement} is calculated as
    \begin{equation}
        \notag
        r_i^2(t) = \left| \vec{r}_i(t) - \vec{r}_i(0) \right|^2
    \end{equation}
    for atom $i$. Solving this for $D$ gives yields
    \begin{equation}
        \notag
        D = \frac{\langle r^2(t) \rangle}{6t}.
    \end{equation}
    We can now make set of simulations for varying initial temperature $T_i$
    and store the final value for the diffusion constant for each final
    temperature $T$. The results are shown in figure
    \ref{fig:diffusion_constant} and tells us that the system melts around a
    temperature of $T = 0.5T_i$, which here amounts to a temperature of about
    $304K$.

    \begin{figure}[h]
        \centering
        \includegraphics[width=0.8\linewidth]{diffusion_constant.pdf}
        \caption[Diffusion constant]{The diffusion constant as a function of temperature. We here
                see that $D$ abruptly increases at the temperature of $T = 304K$
                which indicates that the crystal is melting.  The two lines are the
                least square solutions to the noisy data before and after,
                respectively. The diffusion constant is given in units of area per second.}
        \label{fig:diffusion_constant}
    \end{figure}
\clearpage
